---
title: "Zero Shot and GO analysis"
author: "Timothy Mulderrig"
date: "2026-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install Packages
```{r}
library(vcfR) #install.packages("vcfR")
library(here) #install.packages("here")
library(ggplot2) #install.packages("ggplot2")
library(httr) #install.packages("httr")
library(jsonlite) #install.packages("jsonlite")
library(xml2) #install.packages("xml2")
library(readr) #install.packages("readr")
library(devtools) #install.packages("devtools")
install_github("TriticeaeToolbox/BrAPI.R")
library(BrAPI)
library(R.utils) #install.packages("R.utils")
install.packages("BiocManager")
BiocManager::install("biomaRt")
library(biomaRt)
```
## Load Plant CAD Zero-Shot Output
```{r}
zero_shot <- vcfR::read.vcfR(here::here("zero_shot_4A.vcf"))
```

Now we need to extract the zero shot score estimates for each marker

```{r}
marker_info <- as.data.frame(zero_shot@fix) |> 
  dplyr::mutate(zero_shot_score = as.numeric(sub(".*=","",INFO)))

marker_info$VIDA <- filtered_vcf@gt[,"VIDA"]
marker_info$YELLOWSTONE <- filtered_vcf@gt[,"YELLOWSTONE"]
```

Let's plot the distribution of zero-shot score estimates
```{r}
ggplot(data = marker_info)+
  geom_histogram(aes(x=zero_shot_score), fill="darkgreen", color="black", bins = 30)+
  labs(title="Distribution of Zero-Shot Scores of Chromosome 4A QTL Region SNPs", x="Zero-Shot Score", y= "Count")+
  geom_vline(xintercept = quantile(marker_info$zero_shot_score,0.1), color="black", linetype= "dashed")+
  annotate(
    "text",
    x = quantile(marker_info$zero_shot_score,0.1),
    y = 750,             # adjust based on your histogram counts
    label = stringr::str_wrap("10th percentile of zero-shot scores: -3.25",width = 50),
    color = "black",
    size = 4
  )

ggsave(here::here("zero_shot_score_dist.png"))

marker_info$POS <- as.numeric(marker_info$POS)
library(scales)
ggplot(data = marker_info, aes(x=POS, y= zero_shot_score))+
  geom_point(color="darkgreen")+
  geom_hline(yintercept = 0, color="darkgrey")+
  scale_x_continuous(
    breaks = seq(620000000, 680000000, by = 10000000), 
    labels = comma
  )+
  labs(title="Zero-Shot Scores of SNPs on Chromosome 4A", x="Position (bp)", y="Zero-shot Score")

ggsave(here::here("zero_shot_scores_across_region.png"))

```

Now let's see what genes these SNPs are located nearby, especially for the SNPs with very negative zero-shot scores

```{r functions to pull gene info}
library(httr)
library(readr)

# Function to parse gene IDs and positions from GFF content
parse_gff_content <- function(gff_content) {
  lines <- strsplit(gff_content, "\n")[[1]]
  
  # Keep only lines describing genes
  gene_lines <- grep("\tgene\t", lines, value = TRUE)
  if (length(gene_lines) == 0) return(NULL)
  
  # Extract gene IDs from attributes column
  gene_id <- sub(".*?ID=gene:(.*?);.*", "\\1", gene_lines)
  
  # Extract start and end positions (columns 4 and 5 in GFF)
  gene_start <- as.numeric(sapply(strsplit(gene_lines, "\t"), `[`, 4))
  gene_end   <- as.numeric(sapply(strsplit(gene_lines, "\t"), `[`, 5))
  
  data.frame(
    candidate_gene = gene_id,
    gene_start = gene_start,
    gene_end   = gene_end,
    stringsAsFactors = FALSE
  )
}

extract_genes_from_regions <- function(region_df, species, output_file, chunk_size = 1e6) {
  
  server <- "https://rest.ensembl.org"
  
  all_genes <- data.frame(
    chrom = character(),
    region_start = integer(),
    region_end   = integer(),
    candidate_gene = character(),
    gene_start = integer(),
    gene_end   = integer(),
    stringsAsFactors = FALSE
  )
  
  region_df$start_bp <- as.numeric(region_df$start_bp)
  region_df$end_bp   <- as.numeric(region_df$end_bp)
  
  for (i in seq_len(nrow(region_df))) {
    chrom <- region_df$chrom[i]
    start_bp <- region_df$start_bp[i]
    end_bp   <- region_df$end_bp[i]
    
    # Split into chunks
    chunk_starts <- seq(start_bp, end_bp, by = chunk_size)
    chunk_ends   <- pmin(chunk_starts + chunk_size - 1, end_bp)
    
    for (j in seq_along(chunk_starts)) {
      query_start <- chunk_starts[j]
      query_end   <- chunk_ends[j]
      
      ext <- paste0("/overlap/region/", species, "/", chrom, ":",
                    sprintf("%d", query_start), "-",
                    sprintf("%d", query_end),
                    "?feature=gene")
      url <- paste0(server, ext)
      message("Querying: ", url)
      
      res <- GET(url, content_type("text/x-gff3"))
      
      if (status_code(res) != 200) {
        warning("Skipping chunk: ", url, " (status ", status_code(res), ")")
        next
      }
      
      gff_content <- content(res, "text")
      genes_df <- parse_gff_content(gff_content)
      
      # Only proceed if genes_df is not NULL and has rows
      if (!is.null(genes_df) && nrow(genes_df) > 0) {
        genes_df$chrom <- chrom
        genes_df$region_start <- query_start
        genes_df$region_end   <- query_end
        
        all_genes <- rbind(all_genes, genes_df[, c("chrom", "region_start", "region_end",
                                                   "candidate_gene", "gene_start", "gene_end")])
      }
    }
  }
  
  # Remove duplicate genes
  all_genes <- all_genes[!duplicated(all_genes$candidate_gene), ]
  
  write_csv(all_genes, output_file)
  message("Saved ", nrow(all_genes), " genes to ", output_file)
}
```

```{r}
region_df <- data.frame(
  chrom = "chr4A",
  start_bp = min(marker_info$POS),
  end_bp   = max(marker_info$POS)
)

extract_genes_from_regions(region_df, 
                           species = "Triticum_aestivum", 
                           output_file = here::here("region_genes_4A.csv"))

```

```{r}
region_genes_4A <- readr::read_csv(here::here("region_genes_4A.csv"))
mart <- useMart("plants_mart", dataset="taestivum_eg_gene", host="https://plants.ensembl.org")
genes_of_interest <- unique(region_genes_4A$candidate_gene)
go_data <- getBM(attributes=c("ensembl_gene_id","start_position","end_position","go_id","name_1006","namespace_1003"), filters = c("ensembl_gene_id"), values = c(genes_of_interest) ,mart=mart) 

go_data <- go_data |> 
  dplyr::filter(nchar(go_id)>0)

get_go_definitions <- function(go_ids) {
  go_term_defs <- data.frame(NULL)
  
  for (go_id in go_ids){  
    server <- "https://rest.ensembl.org/ontology/id/"

    result <- tryCatch({
      r <- GET(paste(server, go_id, sep = ""), 
               content_type("application/json"))

      stop_for_status(r)

      res <- fromJSON(toJSON(content(r)))
      data.frame(
        go_id = res$accession, 
        go_name = res$name, 
        go_definition = res$definition
      )
      
    }, error = function(e) {
      message(paste("Skipping:", go_id))
      return(NULL)
    })
    
    # only append if successful
    if (!is.null(result)) {
      go_term_defs <- rbind(go_term_defs, result)
    }
  }
  
  return(go_term_defs)
}

go_defs <- get_go_definitions(unique(go_data$go_id))

go_data_full <- go_data |> 
  dplyr::left_join(go_defs, by = "go_id") 
```

Now we just need to append the GO data to our SNP data with Zero-Shot scores by alligning SNP position to gene start and stop position


```{r}
go_data_for_parent_snps <-  marker_info |> 
  dplyr::left_join(
    go_data_full,
    by = dplyr::join_by(
      POS >= start_position,
      POS <= end_position
    )
  ) |> 
  dplyr::select(CHROM,POS,ID,REF,ALT,VIDA,YELLOWSTONE,INFO,zero_shot_score,ensembl_gene_id,start_position,end_position,go_id,name_1006,namespace_1003,go_name,go_definition)

write.csv(go_data_for_parent_snps, here::here("parent_SNP_GOterms_ZeroShot.csv"), row.names = FALSE)
```

